Estimating duration using survival analysis

### 2024-04-01 cis-imperfect-testing.tex
proof 5.3 and 6.3

I start, in \cref{imperf-test:sec:notation}, by introducing some additional notation needed.
I then modify the components of the likelihood in \cref{imperf-test:sec:modifying-p_ia} and \cref{imperf-test:sec:modifying-p_iu} to incorporate false negatives.

@@

$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
\begin{enumerate}
  \item $\vec{\nu}_k \in \set{E}$.
  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
\end{enumerate}
The final two conditions are due the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.

>>>>

I start, in \cref{imperf-test:sec:notation}, by introducing some additional notation needed.
I then modify the components of the likelihood in \cref{imperf-test:sec:modifying-p_ia,imperf-test:sec:modifying-p_iu} to incorporate false negatives.

@@

$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
\begin{enumerate}
  \item $\vec{\nu}_k \in \set{E}$.
  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
\end{enumerate}
The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.

### 2024-04-01 cis-perf-new-ll.tex
proof
proof 5.3 and 6.3

\chapter{Additional details of derivations in \cref{E-perf-test} and \cref{E-imperf-test}}

\section{Defining possible observations} \label{perf-test:sec:conditions-nu-E}

Let $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$ be an arbitrary vector of possible observations, $\vec{\nu} \in \set{E}$ if and only if all the following conditions hold.
\begin{enumerate}
  \item $l^{(b)} 1, r^{(b)}, l^{(e)}, r^{e} + 1 \in \sched_{i}$, that is individual ${i}$ was tested at all these times.
  \item $l^{(b)} \leq r^{(b)} \leq l^{(e)} \leq r^{(e)}$.
  \item $r^{(b)} = \min\{t \in \sched_{i} \ssep t \geq l^{(b)}\}$, that is $r^{(b)}$ is the first time at or after $l^{(b)}$ that individual ${i}$ was tested.
  \item $r^{(e)} = \min\{t \in \sched_{i} \ssep t > l^{(e)}\} - 1$, that is $r^{(e)}$ is the day before the next time after $l^{(e)}$ that individual ${i}$ was tested.
  \item $1 \leq r^{(b)} \leq T$, that is the episode begins in the period of interest.
\end{enumerate}

\subsection{Imperfect testing}

>>>>

\chapter{Defining possible observations} \label{perf-test:sec:conditions-nu-E}

Let $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{(e)}, i]^T$ be an arbitrary vector of possible observations, $\vec{\nu} \in \set{E}$ if and only if all the following conditions hold.
\begin{enumerate}
  \item $l^{(b)} 1, r^{(b)}, l^{(e)}, r^{(e)} + 1 \in \sched_{i}$, that is individual ${i}$ was tested at all these times.
  \item $l^{(b)} \leq r^{(b)} \leq l^{(e)} \leq r^{(e)}$.
  \item $r^{(b)} = \min\{t \in \sched_{i} \ssep t \geq l^{(b)}\}$, that is $r^{(b)}$ is the first time at or after $l^{(b)}$ that individual ${i}$ was tested.
  \item $r^{(e)} = \min\{t \in \sched_{i} \ssep t > l^{(e)}\} - 1$, that is $r^{(e)}$ is the day before the next time after $l^{(e)}$ that individual ${i}$ was tested.
  \item $1 \leq r^{(b)} \leq T$, that is the episode begins in the period of interest.
\end{enumerate}

### 2024-04-01 cis-perfect-testing.tex
proof 5.3 and 6.3

If $j$ is detected, then $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)}$ and $r_j^{(b)}$ are the earliest and latest time the episode could have begun respectively; and $l_j^{(e)}$ and $r_j^{(e)}$ are the earliest and latest time the episode could have ended respectively.

@@

Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection.
Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.

@@

Then:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \nu_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

The relevant information from the CIS data is fully contained in the vector $\na$.
Therefore, the posterior of interest is:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
&= p(\vec\theta) \sum_{\ntot= \ndet}^{\infty} p(\ntot, \na \mid \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{k \in \set{D}} p_k &\text{by \cref{perf-test:eq:multinomial}} \\
&= p(\vec{\theta}) \left( \prod_{k \in \set{D}} p_k \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right)
\label{perf-test:eq:posterior1}
\intertext{For convenience, define the summation term as:}
\eta &= 
\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
\label{perf-test:eq:eta}
\end{align}
The rest of this section derives expressions for each of $p_{k}$, $p_{u}$ and $\eta$.

@@

Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$.

@@

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{e}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(l)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
Recall the definition $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Therefore:
\begin{align}
  1 - p_u
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}))
  \label{perf-test:eq:pu}
\end{align}
assuming $P(i(j) = i \mid \vec{\theta}) = 1/\Ncis$ as before.
Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta})$.
Therefore, the crucial component is $1 - p_{iu}$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of the episode being detected.

@@

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
% &= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \min \sched_{i(j)} < B_j \leq T_{i(j)}) \\
&= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
\label{perf-test:eq:piu}
\end{align}

>>>>

If $j$ is detected, then $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)}$ and $r_j^{(b)}$ are the earliest and latest time the episode could have begun respectively; and $l_j^{(e)}$ and $r_j^{(e)}$ are the earliest and latest time the episode could have ended respectively.

@@

Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection.
Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.

@@

Then:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \vec{\nu}_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

The relevant information from the CIS data is fully contained in the vector $\na$.
Therefore, the posterior of interest is:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
&= p(\vec\theta) \sum_{\ntot= \ndet}^{\infty} p(\ntot, \na \mid \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{k \in \set{D}} p_k &\text{by \cref{perf-test:eq:multinomial}} \\
&= p(\vec{\theta}) \left( \prod_{k \in \set{D}} p_k \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right).
\label{perf-test:eq:posterior1}
\intertext{For convenience, define the summation term as:}
\eta &= 
\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
\label{perf-test:eq:eta}
\end{align}
The rest of this section derives expressions for each of $p_{k}$, $p_{u}$ and $\eta$.

@@

Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$.

@@

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{(e)}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
Recall the definition $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Therefore:
\begin{align}
  1 - p_u
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}))
  \label{perf-test:eq:pu}
\end{align}
assuming $P(i(j) = i \mid \vec{\theta}) = 1/\Ncis$ as before.
Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta})$.
Therefore, the crucial component is $1 - p_{iu}$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of the episode being detected.

@@

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
% &= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \min \sched_{i(j)} < B_j \leq T_{i(j)}) \\
&= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
&\propto \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
\label{perf-test:eq:piu}
\end{align}
